\large \textbf{Burrows-Wheeler transform for terabases}

### 2015-11-05 paper/paper.tex
Almost down to the 10-page limit.

\begin{abstract}
In order to avoid the reference bias introduced by mapping reads to a reference genome, bioinformaticians are investigating reference-free methods for analyzing sequenced genomes. With large projects sequencing thousands of individuals, this raises the need for tools capable of processing terabases of sequence data. A key method is the Burrows-Wheeler transform (BWT), which is widely used for compressing and indexing reads. We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of smaller subcollections. With our largest 2.4 Tbp datasets, the algorithm can merge 600 Gbp/day on a single system, using 30 gigabytes of memory overhead on top of the run-length encoded BWTs.
\end{abstract}

@@

\emph{De novo assembly} of sequenced genomes is still too difficult to be routinely done. As a practical alternative, bioinformaticians typically align the reads to a \emph{reference genome} of the same species. Because most reference genomes have been derived from the genomes of a small number of individuals, this introduces \emph{reference bias}, which may adversely affect the results of subsequent analysis. Switching from reference sequences to \emph{reference graphs} can help to reduce the bias, but such transition will likely take years \cite{Church2015}.

@@

Read Server partitions the unique reads based on the last two bases of the read. Every query must be repeated in all 16 indexes, and the \BWT{}s require more space, as the compression cannot take advantage of the similarities between the reads in different indexes. Reducing the number of indexes would improve both memory usage and query performance. However, the lack of \BWT{} construction algorithms that work with terabases of sequence data has made it infeasible so far.

There are four often contradictory requirements for large-scale \BWT{} construction:
\begin{itemize}

\item \textbf{Speed.} Larger datasets require faster algorithms. As a rough guideline, an algorithm processing 1~Mbp/s is good for up to 100~Gbp, while remaining somewhat useful until 1~Tbp of data.

\item \textbf{Memory.} We may have to process $n$~bp datasets on systems with less than $n$~bits of memory.

\item \textbf{Hardware.} A single node in a computer cluster typically has tens of CPU cores, from tens to hundreds of gigabytes of memory, a limited amount of local disk space, and access to a large amount of shared disk space with no performance guarantees. Algorithms using a GPU or a large amount of fast disk space require special-purpose hardware.

\item \textbf{Efficiency.} Large \BWT{}s can be built by doing a lot of redundant work on multiple nodes. As most computer clusters do not have large amounts of unused capacity, such inefficient algorithms are not suitable for repeated use.

\end{itemize}

The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} for the sequences using a fast general-purpose algorithm \cite{Mori2008,Nong2011}. Given the sequences and their suffix array, we can easily build the \BWT{} of the sequences. These algorithms cannot be used with large datasets, as they require several times more memory than the sequences themselves. Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index the data faster than 1\nobreakdash--2~Mbp/s \cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Kaerkkaeinen2015a}.

There are several algorithms for constructing the \BWT{} \emph{directly} without having to build the suffix array first. They may require only a limited amount of working space on top of a plain or compressed representation of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}, or use the disk as additional working space \cite{Ferragina2012,Beller2013}. General-purpose implementations of these algorithms rarely exceed 1\nobreakdash--2~Mbp/s with large datasets.

\emph{Specialized algorithms} for DNA sequences achieve better time/space trade-offs. Some algorithms can index 5\nobreakdash--10~Mbps on general-purpose hardware, with their memory usage becoming a major bottleneck for datasets larger than about 1~Tbp \cite{Bauer2013,Li2014a}. GPU-based algorithms can be much faster, but their memory usage is also higher \cite{Liu2014,Pantaleoni2014}. Distributing the \BWT{} construction on multiple nodes can remove the obvious bottlenecks, but such algorithms generally do not use the hardware as efficiently as those running on  a single system \cite{Wang2015}.

In this paper, we propose a practical algorithm for building the \BWT{} for terabases of sequence data. The algorithm is based on dividing the sequence collection into a number of smaller subcollections, building the \BWT{} for each subcollection, and \emph{merging} the \BWT{}s into a single structure \cite{Siren2009}. The merging algorithm is faster than \BWT{} construction for the subcollections, while having a relatively small memory overhead on top of the final \BWT-based index. As the index must be loaded in memory before it can be used, index construction can be done on the same system as the index is going to be used.

@@

The \emph{suffix array} (\SA) \cite{Manber1993} is a simple full-text index of a string. Given a text $T$, its suffix array $\mSA_{T}[1,n]$ is an array of pointers to the suffixes of the text in \emph{lexicographic order}. (If the text is evident from the context, we will omit it and write just \SA.) We can build the suffix array in $\Oh(n)$ time using approximately $n(\log n + \log \sigma + 2)$ bits of space, or just $2n$ bits of working space over the text and the suffix array \cite{Nong2011}. Given a \emph{pattern} $P$, we can binary search the suffix array to find the \emph{lexicographic range} $[sp,ep]$ of suffixes prefixed by the pattern in $\Oh(\abs{P} \log n)$ time. The corresponding range of pointers $\mSA[sp,ep]$ lists the \emph{occurrences} of the pattern in the text.

While the suffix array is simple, it requires several times more memory than the original text. For large texts, this can be a serious drawback. We can use the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994} as a more space-efficient alternative to the suffix array. The \BWT{} is an easily reversible permutation of the text with a similar combinatorial structure to the suffix array. Given a text $T[1,n]$ and its suffix array, we can easily produce the \BWT{} as $\mBWT[i] = T[\mSA[i]-1]$ (with $\mBWT[i] = T[n]$, if $\mSA[i] = 1$).

A key observation is that if suffix $X$ precedes suffix $Y$ in lexicographic order, suffix $cX$ will also precede suffix $cY$, for any character $c$. Therefore, if $\mBWT[i]$ is the $j$\nobreakdash-th occurrence of character $c$ in the \BWT and $\mSA[i]$ points to suffix $X$, suffix $cX$ is the $j$\nobreakdash-th suffix starting with $c$ in lexicographic order. This forms the essence of \LF\emph{\nobreakdash-mapping}. The pointer to suffix $T[\mSA[i]-1,n]$ is stored at $\mSA[\mLF(i)]$, where
$$
\mLF(i) = \mC[\mBWT[i]] + \mBWT.\mrank(i, \mBWT[i]),
$$
$\mC[c]$ is the number of suffixes starting with a character smaller than $c$, and $S.\mrank(i,c)$ is the number of occurrences of character $c$ in the prefix $S[1,i]$. For searching, we use a more general definition of \LF\nobreakdash-mapping:
$$
\mLF(i,c) = \mC[c] + \mBWT.\mrank(i, c),
$$
which tells the number of suffixes $X$ of text $T$ with $X \le c T[\mSA[i],n]$ in lexicographic order. This is known as the \emph{lexicographic rank} $\mrank(cT[\mSA[i],n], T)$ of string $cT[\mSA[i],n]$ among the suffixes of text $T$.

The \emph{FM-index} (\FMI) \cite{Ferragina2005a} uses the Burrows-Wheeler transform as a full-text index by augmenting it with a few additional structures. The lexicographic range $[sp,ep]$ corresponding to pattern $P$ is found through \emph{backward searching}. Let $[sp_{i},ep_{i}]$ be the lexicographic range of suffixes of text $T$ matching suffix $P[i, \abs{P}]$ of the pattern. We can find $[sp_{i-1},ep_{i-1}]$ as
$$
[sp_{i-1},ep_{i-1}] = [\mLF(sp_{i}-1, P[i-1]) + 1, \mLF(ep_{i}, P[i-1])].
$$
By starting from $[sp_{\abs{P}}, ep_{\abs{P}}] = [\mC[P[\abs{P}]]+1, \mC[P[\abs{P}]+1]]$, we can find the lexicographic range of suffixes starting with the pattern in $\Oh(\abs{P} \cdot t_{r})$ time, where $t_{r}$ is the time required to answer \rank{} queries on the \BWT. In practice, the time complexity ranges from $\Oh(\abs{P})$ to $\Oh(\abs{P} \log n)$, depending on the encoding of the \BWT{}.

In order to find the occurrences of the pattern in the text, the FM-index \emph{samples} some suffix array pointers, including the one to the beginning of the text. Unsampled pointers can be derived by using \LF\nobreakdash-mapping. If $\mSA[i]$ is not sampled, the FM-index proceeds to $\mLF(i)$ and continues from there. If $\mSA[\mLF^{k}(i)]$ is the first sample encountered, $\mSA[i] = \mSA[\mLF^{k}(i)] + k$. Depending on the way the samples are selected, we may need a binary sequence to mark the pointers that have been sampled.

Assume that we have an ordered \emph{collection} of texts $\Acoll = (T_{1}, \dotsc, T_{m})$ of total length $n = \abs{\Acoll} = \sum_{i} \abs{T_{i}}$. We want to build a (generalized) \BWT{} for the collection. The usual way is to assume that the endmarkers of all texts are distinct, with the one at the end of text $T_{i}$ having character value $(0,i)$. This guarantees that each suffix has a unique lexicographic rank among the suffixes of the collection. To save space, we still encode each endmarker as a $0$ in the \BWT{}. Because of this, \LF\nobreakdash-mapping does not work with $c = 0$, and we cannot match patterns spanning text boundaries.

When the texts are short (e.g.~reads), there are more space-efficient alternatives to sampling. Because all endmarkers have distinct character values during sorting, we know that $\mSA[i]$ (with $i \le m)$ points to the endmarker of text $T_{i}$. To reach the endmarker, we iterate
$$
\Psi(i) = \mBWT.\mselect(i - \mC[c], c),
$$
where $c$ is the largest character value with $\mC[c] < i$ and $S.\mselect(i,c)$ finds the $i$\nobreakdash-th occurrence of character $c$ in string $S$. If $k \ge 0$ is the smallest value for which $j = \Psi^{k}(i) \le m$, we know that $\mSA[i]$ points to offset $\abs{T_{j}} - k$ in text $T_{j}$.

We can \emph{extract} text $T_{i}$ in $\Oh(\abs{T_{i}} \cdot t_{r})$ time by using \LF\nobreakdash-mapping \cite{Burrows1994}. We start from the endmarker at $\mBWT[i]$ and extract the text backwards as
$$
T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)] \,\, \textrm{for} \,\, 1 \le j < \abs{T_{i}}.
$$
As $\mSA[\LF^{j}(i)]$ points to suffix $T_{i}[\abs{T_{i}}-j, \abs{T_{i}}]$, we also extract the lexicographic ranks of all suffixes of text $T_{i}$ in the process.

@@

The FM-index was introduced as a more space-efficient alternative to the suffix array. If we have to build a suffix array first in order to construct an FM-index, we lose a large part of the benefits, and index construction becomes the overall bottleneck. To overcome this bottleneck, we can use \emph{incremental construction algorithms} that build the FM-index directly. Some of them use an adjustable amount of working space on top of the FM-index, making it possible to index text collections that are too large to fit in memory in an uncompressed form.

Assume that we have built the \BWT{} of text $T$, and we want to \emph{transform} the \BWT{} into that of text $cT$, where $c$ is a character. The transformation \cite{Hon2007} is quite straightforward. We first find the pointer $\mSA[i]$ to the beginning of text $T$ (the position $i$, where $\mBWT[i] = 0$). Then we determine the lexicographic rank $j = \mrank(cT, T) = \mC[c] + \mBWT.\mrank(i, c)$ of text $cT$ among the suffixes of text $T$. Finally we transform the \BWT{} by \emph{replacing} the endmarker at $\mBWT[i]$ with the inserted character $c$, and by \emph{inserting} a new endmarker between $\mBWT[j]$ and $\mBWT[j+1]$.

There are six basic ways of using the transformation in a \BWT{} construction algorithm. Instead of inserting a single character at a time, we may use \emph{batch updates} and transform the \BWT{} of text $T$ directly into that of text $XT$, where $X$ is any string \cite{Hon2007}. We can start with the \BWT{}s of text collections $\Acoll$ and $\Bcoll$, and \emph{merge} them into the \BWT{} of collection $\Acoll \cup \Bcoll$ \cite{Siren2009}. We can also have the \BWT{} of a text collection, and \emph{extend} multiple texts by inserting a new character to the beginning of each of them in a single update \cite{Bauer2013}.

In all three cases, we may use either \emph{static} or \emph{dynamic} \cite{Chan2007} structures for the \BWT. Dynamic representations increase the size of the \BWT{} by a constant factor to facilitate fast updates (around 1.5x for \ropebwtii{} \cite{Li2014a}), while static representations require more space overhead for buffering the updates.

\Section{Merging \BWT{}s of text collections}

\begin{figure}[t!]
\includegraphics[width=\textwidth]{merge_example.pdf}
\caption{Merging the \BWT{}s of texts $R$ and $S$. Rank array \RA{} counts the number of suffixes of $R$ that are lexicographically smaller than or equal to the given suffix of $S$. We fill it by starting with $\mRA[1] = 1$ and iterating $\mRA[\mLF_{S}(i)] = \mLF_{R}(\mRA[i], \mBWT_{S}[i])$. Interleaving bitvector $B_{RS}$ tells whether the source of a character in the merged \BWT{} is in $\mBWT_{R}$ or $\mBWT_{S}$. We build it by setting bits $i+\mRA[i]$ to 1 for all $i$.}\label{fig:merge}
\end{figure}

Assume that we have two text collections $\Acoll$ and $\Bcoll$ of total length $n_{\Acoll}$ and $n_{\Bcoll}$, respectively, and that we have already built the \BWT{}s for them. Assume that we store the \BWT{}s in two-level arrays, where the first level contains pointers $b$\nobreakdash-bit \emph{blocks}. If the \BWT{} takes $x$ bits, the two-level array imposes $\frac{x}{b} \log x + \Oh(b)$ bits of space overhead. With $b = \sqrt{x \log x}$, the overhead becomes $\Oh(\sqrt{x \log x})$ bits.

The algorithm for merging $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ into $\mBWT_{\Acoll \cup \Bcoll}$ has three phases \cite{Siren2009}:
\begin{enumerate}

\item \textbf{Search.} We search for all texts in collection $\Bcoll$ by using $\mBWT_{\Acoll}$, and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of collection $\Bcoll$. This takes $\Oh(n_{\Bcoll} t_{r})$ time. If we do not have collection $\Bcoll$ in plain form, we can extract the texts from $\mBWT_{\Bcoll}$ in the same asymptotic time.

\item \textbf{Sort.} We sort the ranks produced in the search phase to build the \emph{rank array} (\RA) of collection $\Bcoll$ relative to collection $\Acoll$. The rank array is defined as $\mRA_{\Bcoll \mid \Acoll}[i] = \mrank(X, \Acoll)$, where $\mSA_{\Bcoll}[i]$ points to suffix $X$. The array requires $n_{\Bcoll} \log n_{\Acoll}$ bits of space, and we can build it in $\Oh(sort(n_{\Bcoll}, n_{\Acoll}))$ time, where $sort(n, u)$ is the time required to sort $n$ integers from universe $[0,u]$. If we extract the texts from $\mBWT_{\Bcoll}$ in the search phase, we can write the ranks directly into the rank array, making this phase unnecessary.

\item \textbf{Merge.} We interleave $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ according to the rank array. If $\mRA_{\Bcoll \mid \Acoll}[i] = j$, the merged \BWT{} will have $j$ characters from $\mBWT_{\Acoll}$ before $\mBWT_{\Bcoll}[i]$. This phase takes $\Oh(n_{\Acoll} + n_{\Bcoll})$ time. By reusing the blocks of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ for $\mBWT_{\Acoll \cup \Bcoll}$ once we have processed them, we can merge the \BWT{}s almost in-place. The total working space is $\Oh(\sqrt{x \log x})$ bits, where $x$ is the maximum of the sizes of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ in bits.

\end{enumerate}
See Figure~\ref{fig:merge} for an example with two texts. The overall time complexity of \BWT{} merging is $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r})$. The algorithm uses $n_{\Bcoll} \log n_{\Acoll} + \Oh(\sqrt{x \log x})$ bits of working space in addition to the \BWT{}s and the structures required to use them as FM-indexes.

Alternatively, we can build the \emph{interleaving bitvector} $B_{\Acoll \cup \Bcoll}$. This bitvector is a binary sequence of length $n_{\Acoll} + n_{\Bcoll}$ that encodes the same information as the rank array. We set $B_{\Acoll \cup \Bcoll}[i + \mRA_{\Bcoll \mid \Acoll}[i]] = 1$ for $1 \le i \le n_{\Bcoll}$, denoting that $\mSA_{\Acoll \cup \Bcoll}[i + \mRA_{\Bcoll \mid \Acoll}[i]]$ points to a suffix of collection $\Bcoll$. This approach uses $n_{\Acoll} + n_{\Bcoll} + \Oh(\sqrt{x \log x})$ bits of working space.

@@

If we split text collection $\Acoll$ of total length $n$ into $p$ \emph{subcollections} of equal size, we can build $\mBWT_{\Acoll}$ incrementally by merging the \BWT{}s of the subcollections. Merging the subcollections takes $\Oh((p+t_{r})n)$ time and uses (essentially) $\min(n, \frac{n}{p} \log n)$ bits of working space.

When the collection is large, the space overhead of the construction algorithm often determines whether we can build the \BWT{} of the collection. Even if a static encoding of the \BWT{} fits in memory, a dynamic encoding may already be too large. The space overhead from the rank array or the interleaving bitvector (or their equivalents in the other space-efficient algorithms) may also be too much. We can make the rank array fit in memory by increasing the number of subcollections, but that can make the construction too slow.

The key to reducing the space overhead is writing the lexicographic ranks to \emph{disk} in the search phase. If we can perform the sort phase on disk, we just need to scan the rank array once during the merge phase. In practice, as fast disk space is often scarce, we may want to \emph{interleave} the sort phase with the search and merge phases, and to \emph{compress} the lexicographic ranks before writing them to disk. In the following, we describe the key ideas that make large-scale \BWT{} merging fast and space-efficient.

>>>>

\begin{abstract}
In order to avoid the reference bias introduced by mapping reads to a reference genome, bioinformaticians are investigating reference-free methods for analyzing sequenced genomes. With large projects sequencing thousands of individuals, this raises the need for tools capable of handling terabases of sequence data. A key method is the Burrows-Wheeler transform (BWT), which is widely used for compressing and indexing reads. We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of subcollections. With our 2.4 Tbp datasets, the algorithm can merge 600 Gbp/day on a single system, using 30 gigabytes of memory overhead on top of the run-length encoded BWTs.
\end{abstract}

@@

\emph{De novo assembly} of sequenced genomes is still too difficult to be routinely done. As a practical alternative, bioinformaticians usually align the reads to a \emph{reference genome} of the same species. Because most reference genomes come from the genomes of a small number of individuals, this introduces \emph{reference bias}, which may adversely affect the results of subsequent analysis. Switching from reference sequences to \emph{reference graphs} can reduce the bias, but such transition will likely take years \cite{Church2015}.

@@

Read Server partitions the unique reads based on the last two bases of the read. Every query must be repeated in all 16 indexes, and the \BWT{}s require more space, as we cannot compress the similarities between the reads in different indexes. Reducing the number of indexes would improve both memory usage and query performance. However, the lack of \BWT{} construction algorithms that work with terabases of sequence data has made it infeasible so far.

There are four often contradictory requirements for large-scale \BWT{} construction:
\textbf{Speed.} Larger datasets require faster algorithms. As a rough guideline, an algorithm processing 1~Mbp/s is good for up to 100~Gbp, while remaining somewhat useful until 1~Tbp of data.
\textbf{Memory.} We may have to process $n$~bp datasets on systems with less than $n$~bits of memory.
\textbf{Hardware.} A single node in a typical computer cluster has tens of CPU cores, from tens to hundreds of gigabytes of memory, a limited amount of local disk space, and access to shared disk space with no performance guarantees. Algorithms using a GPU or a large amount of fast disk space require special-purpose hardware.
\textbf{Efficiency.} Large \BWT{}s can be built by doing a lot of redundant work on multiple nodes. As most computer clusters do not have large amounts of unused capacity, such inefficient algorithms are not suitable for repeated use.

The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} using a fast general-purpose algorithm \cite{Mori2008,Nong2011}. Given the sequences and their suffix array, we can easily build the \BWT{} of the sequences. These algorithms cannot be used with large datasets, as they require several times more memory than the sequences themselves. Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index the data faster than 1\nobreakdash--2~Mbp/s
% DCC:
\cite{Kaerkkaeinen2015a}.
% arXiv:
%\cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Nong2015,Kaerkkaeinen2015a,Liu2015a}.

There are algorithms for constructing the \BWT{} \emph{directly} without having to build the suffix array first. Some require only a limited amount of working space on top of a plain or compressed representation of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}, or use the disk as additional working space \cite{Ferragina2012,Beller2013}. These general-purpose algorithms rarely exceed 1\nobreakdash--2~Mbp/s with large datasets. \emph{Specialized algorithms} for DNA sequences achieve better time/space trade-offs. Some algorithms can index 5\nobreakdash--10~Mbps using ordinary hardware, with their memory usage becoming a bottleneck for datasets larger than about 1~Tbp \cite{Bauer2013,Li2014a}. GPU-based algorithms can be much faster, but their memory usage is also higher \cite{Liu2014,Pantaleoni2014}. Distributing the \BWT{} construction to multiple nodes can remove the obvious bottlenecks, but such algorithms generally use the hardware less efficiently than those running on a single system \cite{Wang2015}.

In this paper, we propose a practical algorithm for building the \BWT{} for terabases of sequence data. The algorithm is based on dividing the sequence collection into a number of subcollections, building the \BWT{} for each subcollection, and \emph{merging} the \BWT{}s into a single structure \cite{Siren2009}. The merging algorithm is faster than \BWT{} construction for the subcollections, while having a relatively small memory overhead on top of the final \BWT-based index. As the index must be loaded in memory for use, it can be built on the same system as it is going to be used.

@@

The \emph{suffix array} (\SA) \cite{Manber1993} is a simple full-text index. Given a text $T$, its suffix array $\mSA_{T}[1,n]$ is an array of pointers to the suffixes of the text in \emph{lexicographic order}.\footnote{If the text is evident from the context, we will omit the subscript and write just \SA{}, \BWT{}, etc.} We can build the suffix array in $\Oh(n)$ time using $2n$ bits of working space on top of the text and the suffix array \cite{Nong2011}. Given a \emph{pattern} $P$, we can find the \emph{lexicographic range} $[sp,ep]$ of suffixes prefixed by the pattern in $\Oh(\abs{P} \log n)$ time. The range of pointers $\mSA[sp,ep]$ lists the \emph{occurrences} of the pattern in the text.

The suffix array requires several times more memory than the original text. For large texts, this can be a serious drawback. We can use the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994} as a more space-efficient alternative to the suffix array. The \BWT{} is an easily reversible permutation of the text with a similar combinatorial structure to the suffix array. Given a text $T[1,n]$ and its suffix array, we can easily produce the \BWT{} as $\mBWT[i] = T[\mSA[i]-1]$ (with $\mBWT[i] = T[n]$, if $\mSA[i] = 1$).

If suffix $X$ precedes suffix $Y$ in lexicographic order, suffix $cX$ will also precede suffix $cY$, for any character $c$. If $\mBWT[i]$ is the $j$\nobreakdash-th occurrence of character $c$ in the \BWT{} and $\mSA[i]$ points to suffix $X$, suffix $cX$ is the $j$\nobreakdash-th suffix starting with $c$ in lexicographic order. This forms the essence of \LF\emph{\nobreakdash-mapping}. Let $\mC[c]$ bet the number of suffixes starting with a character smaller than $c$, and let $S.\mrank(i,c)$ be the number of occurrences of $c$ in the prefix $S[1,i]$. Then $\mSA[\mLF(i)]$, where
$$
\mLF(i) = \mLF(i,\mBWT[i])
\,\, \textrm{and} \,\,
\mLF(i,c) = \mC[c] + \mBWT.\mrank(i, c),
$$
points to the suffix $T[\mSA[i]-1,n]$. The more general version of \LF-mapping $\mLF(i,c)$ tells the number of suffixes $X$ of text $T$ with $X \le c T[\mSA[i],n]$. This is the \emph{lexicographic rank} $\mrank(cT[\mSA[i],n], T)$ of $cT[\mSA[i],n]$ among the suffixes of text $T$.

The \emph{FM-index} (\FMI) \cite{Ferragina2005a} uses the Burrows-Wheeler transform as a full-text index by augmenting it with a few additional structures. The lexicographic range $[sp,ep]$ corresponding to pattern $P$ is found by using \emph{backward searching}. Let $[sp_{i},ep_{i}]$ be the lexicographic range of suffixes of text $T$ matching suffix $P[i, \abs{P}]$ of the pattern. We find $[sp_{i-1},ep_{i-1}]$ as $[\mLF(sp_{i}-1, P[i-1]) + 1, \mLF(ep_{i}, P[i-1])]$. By starting from $[sp_{\abs{P}}, ep_{\abs{P}}] = [\mC[P[\abs{P}]]+1, \mC[P[\abs{P}]+1]]$, we can find the lexicographic range of suffixes starting with the pattern in $\Oh(\abs{P} \cdot t_{r})$ time, where $t_{r}$ is the time required to answer \rank{} queries on the \BWT. In practice, the time complexity ranges from $\Oh(\abs{P})$ to $\Oh(\abs{P} \log n)$, depending on the encoding of the \BWT{}.

The FM-index \emph{samples} some suffix array pointers, including the one to the beginning of the text. When unsampled pointers are needed, they are derived by using \LF\nobreakdash-mapping. If $\mSA[i]$ is not sampled, the FM-index proceeds to $\mLF(i)$ and continues from there. If $\mSA[\mLF^{k}(i)]$ is the first sample encountered, $\mSA[i] = \mSA[\mLF^{k}(i)] + k$.
% arXiv:
%Depending on the way the samples are selected, we may need a binary sequence to mark the pointers that have been sampled.

Assume that we have an ordered \emph{collection} of texts $\Acoll = (T_{1}, \dotsc, T_{m})$ of total length $n = \abs{\Acoll} = \sum_{i} \abs{T_{i}}$. We want to build a (generalized) \BWT{} for the collection. The usual way is to make all endmarkers distinct, giving the one at the end of text $T_{i}$ character value $(0,i)$. This also makes all suffixes of the collection distinct. To save space, we still encode each endmarker as a $0$ in the \BWT{}. Because of this, \LF\nobreakdash-mapping does not work with $c = 0$, and we cannot match patterns spanning text boundaries.

When the texts are short (e.g.~reads), there are more space-efficient alternatives to sampling. Because all endmarkers have distinct values during sorting, we know that $\mSA[i]$ (with $i \le m)$ points to the end of text $T_{i}$. To find the end, we iterate $\Psi(i) = \mBWT.\mselect(i - \mC[c], c)$, where $c$ is the largest value with $\mC[c] < i$ and $S.\mselect(i,c)$ finds the $i$\nobreakdash-th occurrence of character $c$ in string $S$. If $k \ge 0$ is the smallest value for which $j = \Psi^{k}(i) \le m$, we know that $\mSA[i]$ points to offset $\abs{T_{j}} - k$ in text $T_{j}$.

We can \emph{extract} text $T_{i}$ in $\Oh(\abs{T_{i}} \cdot t_{r})$ time by using \LF\nobreakdash-mapping \cite{Burrows1994}. We start from the endmarker at $\mBWT[i]$ and extract the text backwards as $T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)]$ for $1 \le j < \abs{T_{i}}$. As $\mSA[\LF^{j}(i)]$ points to suffix $T_{i}[\abs{T_{i}}-j, \abs{T_{i}}]$, we also find the lexicographic ranks of all suffixes of text $T_{i}$ in the process.

@@

The FM-index was introduced as a more space-efficient alternative to the suffix array. If we need the suffix array to build the FM-index, a large part of this benefit is lost, and index construction becomes the bottleneck. To overcome the bottleneck, we can use \emph{incremental construction algorithms} that build the FM-index directly. Some of them use an adjustable amount of working space on top of the FM-index, making it possible to index text collections larger than the size of the memory.

Assume that we have built the \BWT{} of text $T$, and we want to \emph{transform} the \BWT{} into that of text $cT$, where $c$ is a character \cite{Hon2007}. We find the pointer $\mSA[i]$ to the beginning of text $T$ (where $\mBWT[i] = 0$). Then we determine the lexicographic rank $j = \mrank(cT, T) = \mC[c] + \mBWT.\mrank(i, c)$ of text $cT$ among the suffixes of text $T$. Finally \emph{replace} the endmarker at $\mBWT[i]$ with the inserted character $c$ and \emph{insert} a new endmarker between $\mBWT[j]$ and $\mBWT[j+1]$.

We can use the transformation for \BWT{} construction in several ways. We can do \emph{batch updates} and transform the \BWT{} of text $T$ into that of text $XT$, where $X$ is a string \cite{Hon2007}. We can start with the \BWT{}s of text collections $\Acoll$ and $\Bcoll$, and \emph{merge} them into the \BWT{} of collection $\Acoll \cup \Bcoll$ \cite{Siren2009}. We can also \emph{extend} multiple texts at once by inserting a new character to the beginning of each of them \cite{Bauer2013}. In all cases, we can use either \emph{static} or \emph{dynamic} \cite{Chan2007} structures for the \BWT. Dynamic representations increase the size of the \BWT{} (e.g.~by around 1.5x in \ropebwtii{} \cite{Li2014a}), while static representations require more space overhead for buffering the updates.

\begin{figure}[t!]
\includegraphics[width=\textwidth]{merge_example.pdf}
\caption{Merging the \BWT{}s of texts $R$ and $S$. Rank array \RA{} counts the number of suffixes of $R$ that are lexicographically smaller than or equal to the given suffix of $S$. We fill it by starting with $\mRA[1] = 1$ and iterating $\mRA[\mLF_{S}(i)] = \mLF_{R}(\mRA[i], \mBWT_{S}[i])$. Interleaving bitvector $B_{RS}$ tells whether the source of a character in the merged \BWT{} is in $\mBWT_{R}$ or $\mBWT_{S}$. We build it by setting bits $i+\mRA[i]$ to 1 for all $i$.}\label{fig:merge}%
\vspace{-6pt}
\end{figure}

Assume that we want to merge the \BWT{}s of two text collections $\Acoll$ and $\Bcoll$ of total length $n_{\Acoll}$ and $n_{\Bcoll}$, respectively \cite{Siren2009}. We store the \BWT{}s in two-level arrays, where the first level contains pointers to $b$\nobreakdash-bit \emph{blocks}. If a \BWT{} takes $x$ bits, the space overhead from the array is $\frac{x}{b} \log x + \Oh(b)$ bits. This becomes $\Oh(\sqrt{x \log x})$ bits with $b = \sqrt{x \log x}$. The merging algorithm has three phases: search, sort, and merge. It uses $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r})$ time and $\min(n_{\Bcoll} \log n_{\Acoll}, n_{\Acoll} + n_{\Bcoll}) + \Oh(\sqrt{x \log x})$ bits of working space in addition to the \BWT{}s and the structures required to use them as FM-indexes. See Figure~\ref{fig:merge} for an example with two texts.

\smallbreak\noindent\textbf{Search.} We search for all texts of collection $\Bcoll$ in $\mBWT_{\Acoll}$, and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of collection $\Bcoll$. This takes $\Oh(n_{\Bcoll} t_{r})$ time. If we do not have collection $\Bcoll$ in plain form, we can extract the texts from $\mBWT_{\Bcoll}$ in the same asymptotic time.

\smallbreak\noindent\textbf{Sort.} We build the \emph{rank array} (\RA) of $\Bcoll$ relative to $\Acoll$ by sorting the ranks. The rank array is defined as $\mRA_{\Bcoll \mid \Acoll}[i] = \mrank(X, \Acoll)$, where $\mSA_{\Bcoll}[i]$ points to suffix $X$. The array requires $n_{\Bcoll} \log n_{\Acoll}$ bits of space, and we can build it in $\Oh(sort(n_{\Bcoll}, n_{\Acoll}))$ time, where $sort(n, u)$ is the time required to sort $n$ integers from universe $[0,u]$. If we extracted the texts from $\mBWT_{\Bcoll}$, we can write the ranks directly into the rank array, making this phase trivial. We can also encode the rank array as a binary sequence $B_{\Acoll \cup \Bcoll}$ of length $n_{\Acoll} + n_{\Bcoll}$. This \emph{interleaving bitvector} is built by setting $B_{\Acoll \cup \Bcoll}[i + \mRA_{\Bcoll \mid \Acoll}[i]] = 1$ for $1 \le i \le n_{\Bcoll}$. If $B_{\Acoll \cup \Bcoll}[j] = 1$, we know that $\mSA_{\Acoll \cup \Bcoll}[j]$ points to a suffix of $\Bcoll$.

\smallbreak\noindent\textbf{Merge.} We interleave $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ according to the rank array. If $\mRA_{\Bcoll \mid \Acoll}[i] = j$, the merged \BWT{} will have $j$ characters from $\mBWT_{\Acoll}$ before $\mBWT_{\Bcoll}[i]$. This phase takes $\Oh(n_{\Acoll} + n_{\Bcoll})$ time. By reusing the blocks of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ for $\mBWT_{\Acoll \cup \Bcoll}$, we can merge the \BWT{}s almost in-place. The total working space is $\Oh(\sqrt{x \log x})$ bits, where $x$ is the maximum of the sizes of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ in bits.

@@

If we split a text collection $\Acoll$ of total length $n$ into $p$ \emph{subcollections} of equal size, we can build $\mBWT_{\Acoll}$ incrementally by merging the \BWT{}s of the subcollections. This takes $\Oh((p+t_{r})n)$ time and uses essentially $\min(n, \frac{n}{p} \log n)$ bits of working space.

When the collection is large, the space overhead of the construction algorithm often determines whether we can build the \BWT{} of the collection.
% arXiv:
%Even if a static encoding of the \BWT{} fits in memory, a dynamic encoding may already be too large. The space overhead from the rank array or the interleaving bitvector (or their equivalents in the other space-efficient algorithms) may also be too much. We can make the rank array fit in memory by increasing the number of subcollections, but that can make the construction too slow.
%
We can reduce the overhead by writing the lexicographic ranks to \emph{disk} in the search phase. If we can sort the ranks on disk, we just need to scan the rank array once during the merge phase. As fast disk space is often scarce, we may want to \emph{interleave} the sorting with the search and merge phases, and to \emph{compress} the ranks before writing them to disk. We now describe the key ideas that make \BWT{} merging fast and space-efficient.
